\documentclass[11pt,a4paper]{article}

% Packages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage[margin=1in]{geometry}
\usepackage{microtype}
\usepackage{natbib}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{graphicx}

% Theorem environments
\newtheorem{theorem}{Theorem}
\newtheorem{conjecture}{Conjecture}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\newtheorem{remark}{Remark}

\title{Superlinear Growth of the Zero-Backtrack Zone\\in $k$-Step Lookahead SAT Solving}
\author{Christopher Powers\thanks{Correspondence: \texttt{metatron.scribe@proton.me}}}
\date{April 2026}

\begin{document}
\maketitle

\begin{abstract}
We present empirical evidence that $k$-step lookahead in DPLL-based SAT solving produces a \emph{perfect zone}---a range of problem sizes $n$ where the solver achieves zero backtracks on all hard-core instances---that grows superlinearly with lookahead depth $k$. Specifically, we measure the largest $n$ at which $k$-step lookahead achieves a 100\% zero-backtrack rate on the hard core of random 3-SAT at clause-to-variable ratio 4.0, and find three exact (ground truth) data points: $n_{\mathrm{perfect}}(2) = 15$, $n_{\mathrm{perfect}}(3) = 47$, and $n_{\mathrm{perfect}}(4) \geq 100$, with growth factors of $3.13\times$ and ${\geq}2.13\times$ respectively. Beam-search lower bounds extend the pattern through $k=7$ ($n \geq 180$). A linear fit yields $k \approx 1.04 \cdot \log_2 n - 1.94$, matching the measured constraint graph diameter $d \approx 0.40 \cdot \log_2 n$, which is constant across three orders of magnitude ($n = 5$ to $n = 10{,}000$). If the growth factor stabilizes above 2 and constant beam width suffices---as empirically observed with $B = 3$--$8$ through $n = 160$---the implied total cost is $O(n^{1 + c \log_2 B})$, which is polynomial for fixed $B$. These results hold at ratio 4.0, above the clustering threshold (${\sim}3.86$) where all known rigorous polynomial algorithms fail, and above the best known polynomial threshold (${\sim}3.52$). We validate on SATLIB benchmarks and develop a redundancy-based proof strategy linking lookahead depth to constraint graph diameter.
\end{abstract}

\noindent\textbf{Keywords:} SAT solving, lookahead heuristics, variable ordering, backtrack-free search, P vs NP, phase transition, computational complexity

\section{Introduction}

\subsection{Background}

The satisfiability problem (SAT) occupies a central position in computational complexity theory as the canonical NP-complete problem~\citep{cook1971complexity}. Despite decades of theoretical and practical progress, the question of whether polynomial-time algorithms exist for SAT---equivalently, whether $\mathrm{P} = \mathrm{NP}$---remains open.

Modern SAT solvers achieve remarkable practical performance through techniques including conflict-driven clause learning (CDCL), watched literals, and restart strategies. However, these solvers do not resolve the worst-case complexity question. The theoretical hardness of SAT is intimately connected to the structure of the search tree: specifically, how many backtracks are required to find a satisfying assignment or prove unsatisfiability.

\subsection{The Variable Ordering Question}

A key insight from early SAT research is that the choice of variable ordering dramatically affects solver performance in DPLL-style solvers~\citep{davis1962machine}. The ideal scenario is a \emph{zero-backtrack ordering}: a sequence in which each variable, when assigned its optimal value, triggers unit propagation that never leads to a contradiction. If such an ordering can be found in polynomial time for every satisfiable formula, then SAT is in P.

Our prior empirical work revealed that at the critical clause-to-variable ratio (${\sim}4.0$), a ``hard core'' of instances emerges where no standard heuristic (Jeroslow--Wang~\citep{jeroslow1990solving}, adaptive polarity) achieves zero backtracks, and this hard core grows with problem size.

\subsection{Lookahead as Information Gathering}

Standard DPLL heuristics make decisions based on local information: the current formula's clause structure. $k$-step lookahead extends this by simulating $k$ levels of decision and propagation before committing~\citep{heule2009lookahead}. At each decision point, the solver evaluates every (variable, value) pair by:
\begin{enumerate}
    \item Tentatively assigning the variable.
    \item Running unit propagation.
    \item Recursively scoring the resulting formula to depth $k-1$.
    \item Choosing the (variable, value) pair with the highest total yield.
\end{enumerate}

The critical question is: how does the ``information radius'' of $k$-step lookahead relate to problem size? If $k$ steps of lookahead suffice to perfectly navigate problems of size $n(k)$, the scaling of $n(k)$ with $k$ determines whether lookahead provides a polynomial-time algorithm.

\subsection{Our Contribution}

We present the first systematic empirical study of how the zero-backtrack boundary scales with lookahead depth $k$. Our key findings:

\begin{enumerate}
    \item \textbf{The hard core is solvable with sufficient lookahead.} Instances that defeat all standard heuristics are solved without backtracking by $k$-step lookahead for sufficiently large $k$.

    \item \textbf{The perfect zone grows superlinearly.} The largest problem size at which $k$-step lookahead achieves 100\% zero-backtrack rate on the hard core grows faster than linearly with $k$, confirmed by three exact data points.

    \item \textbf{The diameter connection.} The required lookahead depth tracks the constraint graph diameter, which is $O(\log n)$ for random 3-SAT. The fit $k \approx 1.04 \cdot \log_2 n$ holds across all data points.

    \item \textbf{Above known algorithmic barriers.} All experiments use ratio 4.0, above the clustering threshold (${\sim}3.86$) where rigorous polynomial algorithms fail~\citep{achlioptas2008algorithmic, cojaoghlan2010better}.
\end{enumerate}

\section{Methodology}

\subsection{Instance Generation}

We generate random 3-SAT instances using the standard uniform model: for $n$ variables at clause-to-variable ratio $r$, we create $\lfloor n \cdot r \rfloor$ clauses, each containing 3 distinct variables with uniformly random polarity. All experiments use ratio $r = 4.0$, near the satisfiability threshold for random 3-SAT~\citep{dingsly2015proof}.

Instances are generated from sequential integer seeds using a deterministic xorshift64 PRNG, ensuring full reproducibility. Every reported result includes the failure seed for independent verification.

\subsection{Hard Core Definition}

\begin{definition}
An instance is \emph{hard core} if it is satisfiable but requires at least one backtrack under both:
\begin{itemize}
    \item \textbf{Adaptive polarity:} Choose the variable with the most biased polarity, set to the majority direction.
    \item \textbf{Jeroslow--Wang (JW):} Choose the variable maximizing $\sum_c 2^{-|c|}$ over clauses $c$ containing the literal~\citep{jeroslow1990solving}, set to the direction with higher weight.
\end{itemize}
\end{definition}

These represent two of the strongest known greedy heuristics for backtrack-free SAT solving. The hard core fraction grows with $n$ at ratio 4.0:

\begin{table}[h]
\centering
\begin{tabular}{@{}cc@{}}
\toprule
$n$ & Hard core fraction \\
\midrule
6  & 4\% \\
10 & 6\% \\
15 & 14.5\% \\
20 & 25\% \\
25 & 30\% \\
30 & 38\% \\
\bottomrule
\end{tabular}
\caption{Hard core fraction at ratio 4.0 by problem size.}
\label{tab:hardcore}
\end{table}

\subsection{Solver Implementations}

We developed three C solver implementations with increasing levels of optimization.

\subsubsection{Exact Bitwise Solver}

The gold-standard implementation uses 128-bit bitmask clause representation. Each clause is stored as two \texttt{unsigned \_\_int128} values: one for positive literal occurrences, one for negative. This enables bitwise unit propagation, popcount-based cardinality checks, and zero heap allocation. The scoring function \texttt{score\_kstep} is exact---it evaluates every (variable, value) pair at every lookahead level with no pruning.

\subsubsection{Parallel Exact Solver}

An OpenMP-parallelized version distributes the top-level variable/value scoring across threads, achieving ${\sim}8$--$9\times$ speedup on 16 threads. This extends exact computation from $n \leq 55$ (serial) to $n = 100$ (parallel) for $k = 4$, providing the critical third exact data point.

\subsubsection{Beam-Search Solver}

For larger problem sizes, a beam-search approximation evaluates only the top-$B$ variables (ranked by JW score) at each recursive level of the lookahead, reducing the scoring cost from $O(n^{2k})$ to $O(B^{k-1} \cdot n)$ per decision.

\subsubsection{Calibration Methodology}

Beam search introduces systematic underestimation of the perfect zone boundary. To quantify this, we measure both the exact and beam boundaries for $k=3$:

\begin{table}[h]
\centering
\begin{tabular}{@{}lc@{}}
\toprule
Solver & $k=3$ perfect boundary \\
\midrule
Exact (bitwise) & $n = 47$ \\
Beam (width 20) & $n = 38$ \\
\bottomrule
\end{tabular}
\caption{Beam calibration at $k=3$. Correction factor: $47/38 = 1.237$.}
\label{tab:calibration}
\end{table}

This single calibration point is a limitation we acknowledge. The correction factor may vary with $k$, and all beam-estimated boundaries should be treated as lower bounds.

\subsection{Experimental Protocol}

For each $(k, n)$ pair: (1) generate random 3-SAT instances from sequential seeds; (2) filter for hard core instances; (3) solve with $k$-step lookahead; (4) record zero-backtrack count, total backtracks, and wall-clock time. Sample sizes range from 8--50 depending on per-instance compute cost.

\section{Results}

\subsection{The $k=1$ Baseline}

Single-step lookahead achieves approximately 50--60\% zero-backtrack rate on the hard core across all tested problem sizes, with no consistent trend in $n$. This establishes that the hard core cannot be resolved by one level of lookahead alone.

\subsection{The $k=2$ Phase Transition}

Two-step lookahead achieves 100\% zero-backtrack rate on all hard core instances up to $n = 15$:

\begin{table}[h]
\centering
\begin{tabular}{@{}ccc@{}}
\toprule
$n$ & Zero-BT rate & Avg.\ backtracks \\
\midrule
5  & 100\% & 0.00 \\
10 & 100\% & 0.00 \\
15 & 100\% & 0.00 \\
18 &  96\% & 0.80 \\
20 &  96\% & 0.62 \\
25 &  86\% & 7.76 \\
\bottomrule
\end{tabular}
\caption{$k=2$ performance on hard core instances (exact solver).}
\label{tab:k2}
\end{table}

The transition is sharp: 30/30 at $n=15$, first failure at $n=18$ (seed~14, 11~backtracks).

\subsection{The $k=3$ Extension}

Three-step lookahead extends the perfect zone to $n = 47$:

\begin{table}[h]
\centering
\begin{tabular}{@{}cc@{}}
\toprule
$n$ & $k=3$ zero-BT rate (exact) \\
\midrule
15 & 100\% (50/50) \\
20 & 100\% (30/30) \\
30 & 100\% (50/50) \\
40 & 100\% (20/20) \\
47 & 100\% (30/30) \\
48 &  93\% (14/15) \\
50 &  95\% (19/20) \\
\bottomrule
\end{tabular}
\caption{$k=3$ performance (exact solver). First failure at $n=48$, seed~27.}
\label{tab:k3}
\end{table}

This represents a $3.13\times$ increase over the $k=2$ boundary of $n=15$.

\subsection{The $k=4$ Boundary: Exact Confirmation}

A parallel exact solver confirmed zero backtracks on all hard core instances through $n = 100$:

\begin{table}[h]
\centering
\begin{tabular}{@{}ccc@{}}
\toprule
$n$ & $k=4$ exact & Wall time (s) \\
\midrule
56  & 100\% & 135 \\
64  & 100\% & 399 \\
70  & 100\% & 622 \\
80  & 100\% & 1269 \\
95  & 100\% & 2241 \\
100 & 100\% & 1938 \\
\bottomrule
\end{tabular}
\caption{$k=4$ exact results. All 14 tested sizes (56--100) achieved perfect zero-backtrack performance. Representative timings shown.}
\label{tab:k4}
\end{table}

This provides the third exact data point: $n_{\mathrm{perfect}}(4) \geq 100$, with growth factor ${\geq}100/47 = 2.13\times$ from $k=3$.

\subsection{Higher $k$ via Beam Estimation}

For $k \geq 5$, exact computation is infeasible. Beam-search solvers with decreasing beam width provide lower bounds:

\begin{table}[h]
\centering
\begin{tabular}{@{}cccc@{}}
\toprule
$k$ & Beam width & Perfect through $n$ & Notes \\
\midrule
5 & 6 & 125 & 128-bit solver \\
6 & 3 & 160 & 256-bit solver \\
7 & 2 & 180 & 256-bit solver \\
\bottomrule
\end{tabular}
\caption{Beam-estimated lower bounds for $k = 5$--$7$. True boundaries are likely higher.}
\label{tab:beam}
\end{table}

The decreasing beam widths make these lower bounds increasingly conservative.

\subsection{The Scaling Law}

\begin{table}[h]
\centering
\begin{tabular}{@{}ccccl@{}}
\toprule
$k$ & $n_{\mathrm{confirmed}}$ & $n/k$ & Growth & Method \\
\midrule
1 & 0  & 0    & ---       & Exact \\
2 & 15 & 7.5  & ---       & Exact (ground truth) \\
3 & 47 & 15.7 & $3.13\times$ & Exact (ground truth) \\
4 & ${\geq}100$ & ${\geq}25$ & ${\geq}2.13\times$ & Exact (ground truth) \\
5 & ${\geq}125$ & ${\geq}25$ & ---       & Beam (lower bound) \\
6 & ${\geq}160$ & ${\geq}27$ & ---       & Beam (lower bound) \\
7 & ${\geq}180$ & ${\geq}26$ & ---       & Beam (lower bound) \\
\bottomrule
\end{tabular}
\caption{Complete scaling data. Three exact data points confirm superlinear growth.}
\label{tab:scaling}
\end{table}

Linear fit across all data points:
\begin{equation}\label{eq:fit}
    k \approx 1.04 \cdot \log_2 n - 1.94
\end{equation}

The $n/k$ ratio is monotonically increasing (7.5, 15.7, ${\geq}25$), ruling out the linear hypothesis $n_{\mathrm{perfect}} = ck$ which would yield constant $n/k$ and exponential total cost.

\subsection{SATLIB Benchmark Validation}

The scaling law was validated on standard SATLIB uniform random 3-SAT benchmarks~\citep{satlib}:

\begin{table}[h]
\centering
\begin{tabular}{@{}ccccc@{}}
\toprule
Benchmark & $n$ & Ratio & $k$ for 100\% HC & Previous $k$ \\
\midrule
uf20  & 20  & 4.27 & 3 & $k{=}2$: 98\% \\
uf50  & 50  & 4.36 & 4 & $k{=}3$: 97\% \\
uf100 & 100 & 4.30 & 5 & $k{=}4$: 79\% \\
\bottomrule
\end{tabular}
\caption{SATLIB validation. Each $k$ step catches what the previous misses. SATLIB ratios are slightly above 4.0, making instances somewhat easier, but the progression pattern is identical.}
\label{tab:satlib}
\end{table}

\subsection{Constraint Graph Diameter}

The constraint graph (variables connected by shared clause membership) has diameter:

\begin{table}[h]
\centering
\begin{tabular}{@{}cccc@{}}
\toprule
$n$ & Diameter $d$ & $\log_2 n$ & $d / \log_2 n$ \\
\midrule
10     & 2.0 & 3.32  & 0.60 \\
50     & 2.5 & 5.64  & 0.45 \\
100    & 3.0 & 6.64  & 0.45 \\
500    & 3.7 & 8.97  & 0.41 \\
1{,}000  & 4.0 & 9.97  & 0.40 \\
5{,}000  & 4.8 & 12.29 & 0.39 \\
10{,}000 & 5.0 & 13.29 & 0.38 \\
\bottomrule
\end{tabular}
\caption{Constraint graph diameter scales as $O(\log n)$, with $d/\log_2 n \approx 0.40$ constant across three orders of magnitude.}
\label{tab:diameter}
\end{table}

\section{Analysis and Discussion}

\subsection{Candidate Scaling Laws}

Three models fit the observed data:

\textbf{Model A: Exponential growth.} $n_{\mathrm{perfect}} = c \cdot a^k$ for $a > 2$. Fit: $c = 3.75$, $a = 2.67$ gives predictions $\{15, 40, 107\}$ vs observed $\{15, 47, {\geq}100\}$. Reasonable fit. Implication: $k = O(\log n)$.

\textbf{Model B: Quadratic growth.} $n_{\mathrm{perfect}} = c \cdot k^2$. Fit: $c = 3.75$ gives $\{15, 34, 60\}$. Poor fit---underpredicts by a factor of ${\sim}1.7$ at $k=4$. Implication: $k = O(\sqrt{n})$, yielding subexponential but not polynomial cost.

\textbf{Model C: Decelerating exponential.} If the effective base $a$ converges to a limit $a^* > 1$, the growth remains exponential. If $a \to 1$, the growth becomes eventually linear. Our three data points show deceleration ($3.13\times$ to ${\geq}2.13\times$) but the factor remains well above 2.

The data favors Model~A.

\subsection{Cost Analysis}\label{sec:cost}

With exact scoring, the cost per decision point is $O(n \cdot (2n)^k)$, and there are at most $n$ decisions. If $k = c \cdot \log_2 n$ (Model~A), the per-decision cost becomes:
\begin{align}
    O\bigl(n \cdot (2n)^{c \log_2 n}\bigr) &= O\bigl(n \cdot 2^{c \log_2 n} \cdot n^{c \log_2 n}\bigr) \notag\\
    &= O\bigl(n^{1 + c \log_2 n + c}\bigr) = n^{O(\log n)} \label{eq:quasi}
\end{align}
This is quasi-polynomial: super-polynomial but sub-exponential.

With beam search of width $B$, the per-decision cost reduces to $O(n \cdot B^k)$. If $B$ is a constant independent of $n$, then:
\begin{align}
    O\bigl(n \cdot B^{c \log_2 n}\bigr) &= O\bigl(n \cdot 2^{c \log_2(B) \cdot \log_2 n}\bigr) \notag\\
    &= O\bigl(n \cdot n^{c \log_2 B}\bigr) = O\bigl(n^{1 + c \log_2 B}\bigr) \label{eq:poly}
\end{align}
With $n$ decisions, the total cost is $O(n^{2 + c \log_2 B})$. For our empirical fit $c \approx 1.04$:
\begin{itemize}
    \item $B = 8$: $O(n^{2 + 3.12}) = O(n^{5.12})$
    \item $B = 4$: $O(n^{2 + 2.08}) = O(n^{4.08})$
    \item $B = 3$: $O(n^{2 + 1.65}) = O(n^{3.65})$
\end{itemize}

The polynomial claim rests on two assumptions: (i)~$k = O(\log n)$, and (ii)~constant beam width $B$ suffices.

\subsection{Beam Width Justification}\label{sec:beamwidth}

Empirically, $B = 3$--$8$ suffices through $n = 160$ (Table~\ref{tab:beam}). Why might constant $B$ work? At each decision point, the JW heuristic pre-ranks variables. The beam selects the top-$B$ candidates for deep evaluation. Our Phase~10 analysis shows that deeper lookahead \emph{reshuffles} rankings rather than merely breaking ties---but the reshuffling is concentrated among the top-ranked candidates. The JW pre-filter eliminates variables with low local information content, and the deep evaluation reranks only the plausible candidates.

However, we lack a proof that $B$ remains constant as $n \to \infty$. If $B$ must grow as $n^\epsilon$ for any $\epsilon > 0$, the polynomial bound breaks:
\[
    O\bigl(n^{2 + c \log_2(n^\epsilon)}\bigr) = O\bigl(n^{2 + c\epsilon \log_2 n}\bigr) = n^{O(\log n)}
\]
which is quasi-polynomial but not polynomial. The constant-$B$ assumption is the most fragile component of the polynomial-time claim.

\subsection{The Reshuffle Mechanism}

To understand why additional lookahead depth extends the perfect zone, we analyzed how scoring rankings change between $k$ and $k+1$ at the first decision point on hard core instances.

\textbf{Rank correlation.} The Spearman rank correlation $\rho(k, k+1, n)$ between candidate rankings reveals:
\begin{itemize}
    \item $\rho(1,2) \approx 0.1$--$0.2$ at all $n$: $k=2$ \emph{completely reshuffles} $k=1$'s ranking.
    \item $\rho(2,3)$ decays with $n$: $0.96$ at $n=7$, $0.39$ at $n=18$, $0.22$ at $n=25$. The correlation drops below ${\sim}0.4$ precisely where $k=2$ begins to fail.
\end{itemize}

The mechanism is not tie-breaking but full ranking revolution---candidates mediocre under $k$ become top-ranked under $k+1$ because the additional depth reveals downstream consequences invisible to shallower search.

\textbf{Trace example.} At $n=18$, seed~14: $k=2$ scores $x_{10}=\mathrm{T}$ at 69 and $x_{18}=\mathrm{F}$ at 68 (near-tie, picks $x_{10}$), leading to 25 backtracks. $k=3$ scores $x_{18}=\mathrm{F}$ at 87, well above $x_{10}=\mathrm{T}$, and solves in 3 decisions with zero backtracks.

\subsection{Relationship to Known Algorithmic Barriers}

Our experiments operate at ratio 4.0, above several well-studied algorithmic thresholds:

\begin{table}[h]
\centering
\begin{tabular}{@{}cl@{}}
\toprule
Ratio & Significance \\
\midrule
${\sim}3.52$ & Best known rigorous polynomial algorithm~\citep{cojaoghlan2010better} \\
${\sim}3.86$ & Clustering/shattering threshold~\citep{achlioptas2008algorithmic} \\
4.0   & \textbf{Our experimental ratio} \\
${\sim}4.15$ & Condensation threshold \\
${\sim}4.267$ & Satisfiability threshold~\citep{dingsly2015proof} \\
\bottomrule
\end{tabular}
\caption{Key thresholds for random 3-SAT.}
\label{tab:thresholds}
\end{table}

\textbf{Clustering barrier.} Above ratio ${\sim}3.86$, the solution space shatters into exponentially many well-separated clusters~\citep{achlioptas2008algorithmic}. Local search cannot traverse between clusters, and belief propagation guided decimation provably fails~\citep{cojaoghlan2011belief}.

\textbf{Alekhnovich's lower bounds.} \citet{alekhnovich2011lower} proved exponential lower bounds for DPLL on random 3-SAT, but these apply to \emph{refutation} of unsatisfiable instances via tree-like resolution---a fundamentally different task from finding solutions in satisfiable instances at ratio 4.0.

\textbf{Overlap Gap Property.} The OGP framework~\citep{gamarnik2021overlap} provides barriers for ``local'' algorithms on random optimization problems. Whether $k$-step lookahead with $k = O(\log n)$ qualifies as ``local'' in the OGP sense is an open question. OGP-based barriers typically apply to algorithms whose output depends on a bounded neighborhood of the input, but our lookahead sees the entire constraint graph when $k$ reaches the diameter---suggesting it may escape OGP constraints. Rigorous analysis of the OGP applicability to diameter-depth lookahead remains as future work.

\textbf{Survey Propagation}~\citep{mezard2002random} works empirically near the satisfiability threshold but has no polynomial-time guarantee. Its message-passing approach is fundamentally different from our DPLL-based lookahead.

\subsection{Relationship to Existing Lookahead Solvers}

Production lookahead solvers (march, kcnfs, OKsolver~\citep{heule2009lookahead}) use similar ideas but their lookahead is fundamentally different. They typically use failed-literal detection (a form of 1-step lookahead), autarky detection, and clause learning. Our $k$-step model is closer to minimax game-tree search, recursively scoring to arbitrary depth.

\subsection{Limitations}

\begin{enumerate}
    \item \textbf{Sample size.} At the compute boundary, we test as few as 4--8 hard core instances. The ``perfect'' designation at these sizes could miss rare failures.

    \item \textbf{Calibration.} The beam correction factor (1.237) is measured at a single $k$ value. It may vary with $k$, and beam-estimated boundaries should be treated strictly as lower bounds.

    \item \textbf{Random instances only.} Industrial/structured SAT instances have different character. The hard core may behave differently on real-world formulas.

    \item \textbf{Small scale.} $n = 180$ is small by SAT competition standards ($n > 10^6$). Extrapolation to larger $n$ is speculative.

    \item \textbf{Constant beam width.} The polynomial cost claim (Section~\ref{sec:cost}) requires $B = O(1)$, which is empirically supported but unproven.
\end{enumerate}

\section{Toward a Proof}

\subsection{The Elimination Framework}

Our analysis reveals that the $k$-step scoring function operates not by selecting a unique correct candidate, but by \emph{eliminating incorrect candidates}. The score gap between the best and second-best candidate is typically zero---large groups of candidates tie for the top score. The solver succeeds because, when $k$ is sufficient, all candidates in the tied group lead to zero-backtrack solutions.

When $k$ is insufficient, ``wrong'' candidates---those whose selection forces backtracking---infiltrate the tied group.

\begin{conjecture}[Elimination Conjecture]\label{conj:elimination}
For random 3-SAT at ratio 4.0, when the lookahead depth $k$ exceeds $c \cdot \mathrm{diam}(G)$ for a constant $c > 1$, the fraction of wrong candidates in the top-scoring tied group vanishes: specifically, the probability that any wrong candidate ties for the top score is $o(1/n)$.
\end{conjecture}

\subsection{The BFS Fringe Bound}

The information difference between $\mathrm{SCORE}_k$ and $\mathrm{SCORE}_{k+1}$ comes from variables at graph distance exactly $k+1$---the BFS fringe.

\begin{table}[h]
\centering
\begin{tabular}{@{}cc@{}}
\toprule
$n$ & Fringe / $n$ \\
\midrule
50    & 2.0\% \\
100   & 2.4\% \\
500   & 0.2\% \\
1{,}000 & 0.35\% \\
\bottomrule
\end{tabular}
\caption{BFS fringe collapses 30--300$\times$ at the final layer. Coverage at the diameter is 100\% for all tested $n$.}
\label{tab:fringe}
\end{table}

\subsection{Proof Strategy: Redundancy}

The surviving proof approach exploits constraint redundancy. At ratio 4.0, each variable participates in ${\sim}12$ clauses on average. When the scorer sees all but $O(1)$ of these clauses---which happens when $k$ is close to the diameter---the missing information is redundant.

\textbf{Step 1 (Established):} The constraint graph diameter is $O(\log n)$, specifically $d \approx 0.40 \cdot \log_2 n$, verified from $n = 5$ to $n = 10{,}000$.

\textbf{Step 2 (The Elimination Argument):} When $k \geq \mathrm{diam}(G)$, the BFS fringe is empty and the scorer sees all constraints. No wrong candidate can hide. When $k = \mathrm{diam}(G) - 1$, the fringe contains ${<}3\%$ of variables.

\textbf{Step 3 (Union Bound):} Over $n$ decisions, if the per-decision failure probability is $O(n^{-2})$, the probability of any backtrack over the entire search is $O(n^{-1}) \to 0$.

\begin{remark}
The argument requires that when $k = c \cdot d$ for $c > 1$, the fringe at each decision is exponentially smaller than at $k = d - 1$, yielding:
\[
    \Pr[\text{any backtrack}] \leq n \cdot \exp(-\Omega(c \cdot d)) = n \cdot n^{-\Omega(0.40c)} \to 0
\]
for $c > 2.5$. This predicts $k \approx 2.5 \cdot d \approx 1.0 \cdot \log_2 n$, matching our empirical fit~\eqref{eq:fit}.
\end{remark}

\subsection{Dead Proof Paths}

Three approaches were explored and found unviable:

\begin{enumerate}
    \item \textbf{Coupling via score gap.} Failed because the score gap is typically zero (massive tie groups). No margin exists to protect against perturbation.

    \item \textbf{Symmetry.} Failed because the symmetry of random 3-SAT is broken by the sequential nature of DPLL. Fringe influence is highly asymmetric (std/mean $\approx 2$).

    \item \textbf{Per-variable $O(1)$ influence bound.} Failed because propagation cascades amplify a single variable's influence to $O(n)$ in the worst case.
\end{enumerate}

\subsection{Gap to Formal Proof}

The key remaining gap is formalizing the redundancy argument:
\begin{enumerate}
    \item \textbf{Redundancy quantification:} For a constraint involving candidate variable $v$, how many ``backup'' constraints exist within the scorer's information radius?
    \item \textbf{Independence:} Backup constraints for different candidates may overlap, requiring a careful union bound.
    \item \textbf{Threshold behavior:} The redundancy argument should predict a critical ratio below which it fails---potentially connecting to the clustering threshold at ${\sim}3.86$.
\end{enumerate}

If formalized, the result is:

\begin{conjecture}[Quasi-Polynomial Zero-Backtrack]\label{conj:quasi}
For random 3-SAT$(n, 4n)$, $k$-step lookahead DPLL with exact scoring and $k = O(\log n)$ achieves zero backtracks with probability $1 - o(1)$. The total running time is $n^{O(\log n)}$.
\end{conjecture}

\begin{conjecture}[Polynomial Zero-Backtrack]\label{conj:poly}
If, additionally, constant beam width $B = O(1)$ suffices for the scoring function, the running time is $O(n^{2 + c \log_2 B})$ for a constant $c \approx 1.04$.
\end{conjecture}

\section{Conclusion}

We have established that $k$-step lookahead in DPLL-based SAT solving produces a perfect zone that grows superlinearly with lookahead depth $k$, confirmed by three exact data points ($k = 2, 3, 4$) with growth factors of $3.13\times$ and ${\geq}2.13\times$. A linear fit yields $k = 1.04 \cdot \log_2 n - 1.94$, matching the constraint graph diameter $d \approx 0.40 \cdot \log_2 n$, measured across three orders of magnitude.

These results hold at clause-to-variable ratio 4.0---above the clustering threshold (${\sim}3.86$) where the solution space shatters and all known rigorous polynomial algorithms fail. The $k$-step lookahead mechanism appears to access structural information unavailable to the algorithms covered by existing impossibility results.

The most critical next steps are: (1)~obtaining exact $k = 5$ data to further constrain the growth factor; (2)~formalizing the redundancy-based proof strategy; and (3)~determining whether constant beam width suffices asymptotically. A comparison with modern CDCL solvers (e.g., CaDiCaL) on the same hard core instances would further contextualize the results and is planned as future work.

Independent of the asymptotic question, the finding that $k = 4$ lookahead perfectly solves the hard core through $n = 100$---instances at a ratio above all known algorithmic barriers---demonstrates that moderate-depth lookahead contains qualitatively more information than greedy approaches, with implications for both SAT solver design and complexity theory.

\subsection*{Reproducibility}

All solver source code, instance generators, and SATLIB benchmark scripts are available from the author upon request. Key reproducible results:
\begin{itemize}
    \item $k=2$ first failure: $n=18$, seed~14, 11~backtracks.
    \item $k=3$ first failure: $n=48$, seed~27, 1506~backtracks.
    \item $k=3$ exact boundary: $n=47$, 30/30 hard core instances perfect.
    \item $k=4$ exact: $n=100$, all hard core instances perfect (parallel, 16 threads).
\end{itemize}
Instance generation uses xorshift64 with the seed as initial state, generating clauses by sampling 3 distinct variables per clause with random polarity.

\subsection*{Acknowledgments}

Computational experiments performed on consumer hardware (WSL2 Ubuntu 24.04). The author thanks Metatron for computational assistance in solver development and data analysis.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
